Evolution of pre-copulatory and post-copulatory inbreeding behaviour

### 2016-09-23 ms_pre_post.Rmd
Revise up through Methods

To analyse our model, we track mean values of alleles underlying tendency for polyandry ($P_{a}$), inbreeding mating strategy ($M_{a}$), and inbreeding fertilisation strategy ($F_{a}$) in a focal population. We present evolutionary dynamics of these mean allele values across different parameter combinations to test our four focal hypotheses (see below). We simulate up to eight possible permutations of values of costs of tendency for polyandry ($c_{P}$), inbreeding mating strategy ($c_{M}$), and inbreeding fertilisation strategy ($c_{F}$) equal to either $0$ or $0.02$ [@Duthie observed $0.02$ to be a strong cost of tendency for polyandry] for values of $\beta = 3$ ($8$ total combinations). In all simulations, we record mean values of $P_{a}$, $M_{a}$, $F_{a}$ in each generation. Each combination of parameter values simulated is replicated $40$ times, and grand mean values and standard error of means is calculated in each generation across replicate simulations. These analyses allow us to infer how allele values change over generations in response to costs, but also in response to the changing values of other alleles and therefore potentially interacting phenotypes (e.g., selection on inbreeding fertilisation strategy might be weak if polyandry is rare). For all replicates, we set the maximum number of generations to 10000, which exploratory simulations and previous modelling [@Duthie] showed to be a sufficient number of generations for inferring long-term dynamics of mean allele values and therefore selection on phenotypes. For all replicates, we also set the number of immigrants per generation to $\rho=5$, which produced a range of kin and non-kin in each generation allowing females could express inbreeding strategies. Females always produced $n=8$ offspring, which was sufficient to keep populations consistently near carrying capacities and avoid population extinction. Carrying capacities for females ($K_{f}$) and males ($K_{m}$) were both set to 100 because previous modelling has shown that populations of this size are small enough that mate encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but populations are also not so small that selection is typically overwhelmed by genetic drift (Duthie and Reid *in press*).

To test our hypothesis that selection for evolution of polyandry is driven by a female's ability to bias fertilisation toward inbreeding avoidance through post-copulatory mechanisms, we compare simulations in which all three phenotypes including tendency for polyandry ($P_{p}$), inbreeding mating strategy ($M_{p}$), and inbreeding fertilisation strategy ($F_{p}$) evolve, with an identical set of simulations in which $F_{p}$ cannot evolve. In simulations without evolving $F_{p}$, we therefore include all $4$ combinations of $c_{P}=\{0,0.02\}$ and $c_{M}=\{0,0.02\}$ (but not $c{F}$). To stop $F_{p}$ from evolving, we sever the connection from $F_{a}$ to $F_{p}$ such that all $F_{g}$ genotypes resulting from $F_{a}$ cause random fertilisation with respect to kinship (i.e, $F_{p}=0$). Hence, while mutation and drift can still occur for $F_{a}$ alleles, these alleles have no phenotypic effect and polyandry cannot cause polyandrous females to benefit through an increased ability to bias fertilisation toward inbreeding avoidance (under such conditions, there can also never be a cost of $F_{p}$, meaning that $c_{F}=0$). By comparing evolution of mean $P_{p}$ values over generations when $F_{p}$ also evolves versus when $F_{p}=0$ (random fertilisation), we test whether or not evolution of inbreeding fertilisation strategy increases selection for evolution of polyandry.

To test hypotheses regarding evolution under different relative costs of phenotypes, we consider all combinations of costs using simulations in which all three traits evolve. To test the hypothesis that inbreeding mating strategy will evolve toward inbreeding avoidance ($M_{p}<0$) when mating costs are low relative to the costs of tendency for polyandry and inbreeding fertilisation strategy, but that selection for inbreeding fertilisation strategy will then be negligible ($F_{p} \approx 0$), we quantify change in $M_{a}$ and $F_{a}$ over generations in simulations where $c_{M}=0$ but $c_{P}$, $c_{F}$, or both equals $0.02$. We compare evolution in these three cost parameter value combinations with simulations in which all costs equal either $0$ or $0.02$. If our hypothesis is supported, then $M_{p}$ but not $F_{p}$ will evolve to more negative values when $c_{M}$ is relatively low than when all cost values are uniform. To test the hypothesis that inbreeding fertilisation strategy will evolve toward inbreeding avoidance ($F_{p}<0$) given low costs of polyandry and inbreeding fertilisation strategy relative to the cost of inbreeding mating strategy, but that selection on inbreeding mating strategy will be negligible ($M_{p} \approx 0$), we quantify evolution of $M_{p}$ and $F_{p}$ over generations when $c_{F}=0$, $c_{P}=0$, and $c_{M}=0.02$ . We compare evolution for this combination of cost parameter values with simulations in which all costs equal either $0$ or $0.02$. If our hypothesis is supported, then $F_{a}$ but not $M_{a}$ allele values will become more negative over generations when $c_{F}$ is relatively low than when all cost values are uniform.

To test the hypothesis that constraints on evolution of one inbreeding strategy will increase the strength of selection on a complementary strategy, we fixed inbreeding strategy phenotypes to values of zero, thereby causing either random mating ($M_{p}=0$) or the random fertilisation of offspring ($F_{p}=0$) according to a fair raffle among mates with respect to kinship. We then compare evolution of the non-fixed trait to its evolution in simulations where all traits are allowed to evolve.

To test the hypothesis that the existence of an adaptive inbreeding strategy will decrease the strength of selection on, and thereby preclude the invasion of, a complementary strategy, we first used exploratory simulations to determine evolution of inbreeding mating strategy, and inbreeding fertilisation strategy and associated polyandry, in isolation. To test whether or not inbreeding mating avoidance was predicted to evolve when adaptive polyandry to avoid inbreeding was already established, we initiated inbreeding mating strategy allele values at zero ($M_{a}=0$), but fixed inbreeding fertilisation strategy allele values at $F_{a}=-5$ and polyandry allele values at $P_{a}=1$ (i.e., $F_{p}$ and $P_{p}$ did not evolve). Similarly, to test whether or not inbreeding fertilisation strategy evolved to avoid inbreeding, we initiated inbreeding fertilisation strategy allele values and polyandry allele values at zero ($F_{a}=0$; $P_{a}=0$), but fixed inbreeding mating strategy allele values at $M_{a}=-5$.

>>>>

To analyse our model, we track mean values of alleles underlying tendency for polyandry ($P_{a}$), inbreeding mating strategy ($M_{a}$), and inbreeding fertilisation strategy ($F_{a}$) in a focal population. We present evolutionary dynamics of these mean allele values across different parameter combinations to test our three focal hypotheses.

To test our hypothesis that selection for polyandry is driven by a female's ability to bias fertilisation toward inbreeding avoidance through post-copulatory mechanisms, we compare simulations in which all three phenotypes including tendency for polyandry ($P_{p}$), pre-copulatory inbreeding strategy ($M_{p}$), and post-copulatory strategy ($F_{p}$) evolve, with an identical set of simulations in which $F_{p}$ cannot evolve given $c_{P} = \{0, 0.0025, 0.005}$. To stop $F_{p}$ from evolving, we sever the connection from $F_{a}$ to $F_{p}$ such that all $F_{g}$ genotypes resulting from $F_{a}$ cause random fertilisation with respect to kinship (i.e, $F_{p}=0$). Hence, while mutation and drift can still occur for $F_{a}$ alleles, these alleles have no phenotypic effect and polyandry cannot cause polyandrous females to benefit through an increased ability to bias fertilisation toward inbreeding avoidance. By comparing evolution of mean $P_{p}$ values over generations when $F_{p}$ also evolves versus when $F_{p}=0$ (random fertilisation), we test whether or not evolution of post-copulatory inbreeding strategy increases selection for polyandry.

<!--- We simulate up to eight possible permutations of values of costs of tendency for polyandry ($c_{P}$), inbreeding mating strategy ($c_{M}$), and inbreeding fertilisation strategy ($c_{F}$) equal to either $0$ or $0.02$ [@Duthie observed $0.02$ to be a strong cost of tendency for polyandry] for values of $\beta = 3$ ($8$ total combinations). --->

<!--- In simulations without evolving $F_{p}$, we therefore include all $4$ combinations of $c_{P}=\{0,0.02\}$ and $c_{M}=\{0,0.02\}$ (but not $c{F}$). --->

<!--- TODO: NEED TO DO THESE SIMULATIONS ASAP --->
To test the hypothesis that selection for a costly strategy of inbreeding avoidance will be negligible given a less costly alternative strategy, we quantify change in $M_{a}$ and $F_{a}$ over generations in simulations where the cost of pre-copulatory inbreeding strategy $c_{M}=0$ but the cost of post-copulatory inbreeding strategy $c_{F}=0.02$, and where $c_{F}=0$ but $c_{M}=0.02$. We contrast these simulations with evolution of a costly strategy in the absence of evolution of an alternative strategy (e.g., evolution of pre-copulatory inbreeding strategy when post-copulatory inbreeding strategy phenotype is fixed at zero, $F_{p}=0$). A contrast between the baseline expectation for how costs affect evolution of a focal inbreeding strategy in the absence of an alternative strategy, and evolution of a focal strategy given that an alternative strategy also evolves, tests whether or not a relatively low cost alternative strategy decreases selection for a focal strategy.

<!--- To test hypotheses regarding evolution under different relative costs of phenotypes, we consider all combinations of costs using simulations in which all three traits evolve. To test the hypothesis that inbreeding mating strategy will evolve toward inbreeding avoidance ($M_{p}<0$) when mating costs are low relative to the costs of tendency for polyandry and inbreeding fertilisation strategy, but that selection for inbreeding fertilisation strategy will then be negligible ($F_{p} \approx 0$), we quantify change in $M_{a}$ and $F_{a}$ over generations in simulations where $c_{M}=0$ but $c_{P}$, $c_{F}$, or both equals $0.02$. We compare evolution in these three cost parameter value combinations with simulations in which all costs equal either $0$ or $0.02$. If our hypothesis is supported, then $M_{p}$ but not $F_{p}$ will evolve to more negative values when $c_{M}$ is relatively low than when all cost values are uniform. To test the hypothesis that inbreeding fertilisation strategy will evolve toward inbreeding avoidance ($F_{p}<0$) given low costs of polyandry and inbreeding fertilisation strategy relative to the cost of inbreeding mating strategy, but that selection on inbreeding mating strategy will be negligible ($M_{p} \approx 0$), we quantify evolution of $M_{p}$ and $F_{p}$ over generations when $c_{F}=0$, $c_{P}=0$, and $c_{M}=0.02$ . We compare evolution for this combination of cost parameter values with simulations in which all costs equal either $0$ or $0.02$. If our hypothesis is supported, then $F_{a}$ but not $M_{a}$ allele values will become more negative over generations when $c_{F}$ is relatively low than when all cost values are uniform. --->

<!--- To test the hypothesis that constraints on evolution of one inbreeding strategy will increase the strength of selection on a complementary strategy, we fixed inbreeding strategy phenotypes to values of zero, thereby causing either random mating ($M_{p}=0$) or the random fertilisation of offspring ($F_{p}=0$) according to a fair raffle among mates with respect to kinship. We then compare evolution of the non-fixed trait to its evolution in simulations where all traits are allowed to evolve. --->

To test the hypothesis that the existence of an adaptive inbreeding strategy will decrease the strength of selection on, and thereby preclude the invasion of, a complementary strategy, we first used exploratory simulations to determine evolution of precopulatory inbreeding strategy, and post-copulatory inbreeding strategy and associated polyandry, in isolation. To test whether or not pre-copulatory inbreeding avoidance was predicted to evolve when adaptive polyandry to avoid inbreeding was already established, we initiated pre-copulatory inbreeding strategy allele values at zero ($M_{a}=0$), but fixed post-copulatory inbreeding allele values at $F_{a}=-10$ and polyandry allele values at $P_{a}=1$ (i.e., $F_{p}$ and $P_{p}$ did not evolve). Similarly, to test whether or not inbreeding fertilisation strategy evolved to avoid inbreeding, we initiated inbreeding fertilisation strategy allele values and polyandry allele values at zero ($F_{a}=0$; $P_{a}=0$), but fixed inbreeding mating strategy allele values at $M_{a}=-10$.

In all simulations, we record mean values of $P_{a}$, $M_{a}$, $F_{a}$ in each generation. Each combination of parameter values simulated is replicated $40$ times, and grand mean values and standard error of means is calculated in each generation across replicate simulations. These analyses allow us to infer how allele values change over generations in response to costs, but also in response to the changing values of other alleles and therefore potentially interacting phenotypes (e.g., selection on inbreeding fertilisation strategy might be weak if polyandry is rare). For all replicates, we set the maximum number of generations to 40000, which exploratory simulations and previous modelling [@Duthie] showed to be a sufficient number of generations for inferring long-term dynamics of mean allele values and therefore selection on phenotypes.  For all replicates, we set $\rho=5$ immigrants, which produced a range of kin and non-kin in each generation allowing females could express inbreeding strategies, and $n=8$ offspring, which was sufficient to keep populations consistently near carrying capacities and avoid population extinction. Values of $K_{f}$ and $K_{m}$ were both set to 100 because previous modelling has shown that populations of this size are small enough that mate encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but populations are also not so small that selection is typically overwhelmed by genetic drift (Duthie and Reid *in press*).
